---
title: "Lesson 2 - Graphs and Distributions"
author: "Domingo Salerno"
date: "2023-01-02"
output: html_document
---

# Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r packages}
# Load necessary packages
pacman::p_load(ggplot2, readr)
```

```{r data}
# Load English L2 lexical decision data
lexdec <- read_csv("./data/L2_English_Lexical_Decision_Data.csv",
                   col_names = T,
                   col_types = "cfffiiififfnifff")
```

There was a problem when we loaded our data! The warning message provides us the `problems()` function which we can use to investigate the issue.

Now, let's fix the problem.

```{r fix data}
# Load English L2 lexical decision data
lexdec <- read_csv("./data/L2_English_Lexical_Decision_Data.csv",
                   col_names = T,
                   col_types = "cfffiiifcffnifff")

# Fix "10 years" values
lexdec$monthsLive <- ifelse(lexdec$monthsLive == "10 years",
                            "120",
                            lexdec$monthsLive)

lexdec$monthsLive <- as.integer(lexdec$monthsLive)
```

Examine our data.

```{r}
# Print the structure of the data
str(lexdec)

# Print statistical summary of data
summary(lexdec)

# Check unique values of columns
unique(lexdec$domLang)
```

# Introduction to ggplot2

ggplot2 is a powerful R package for data visualization based on the grammar of graphics concept. It allows for the creation of complex and customizable plots through a layered approach.

Here, we'll create a scatter plot of car mpg vs horsepower using ggplot2.

```{r}
# Create a basic ggplot using the 'mtcars' dataset

# Step 1: Start with the ggplot function
ggplot(data = mtcars, aes(x = hp, y = mpg)) +
  
  # Step 2: Add a geometry layer using the geom_point function for scatter plot
  geom_point() +
  
  # Step 3: Add labels and title using the labs function
  labs(title = "Scatter Plot of MPG vs Horsepower",
       x = "Horsepower",
       y = "Miles per Gallon")
```

The above chunk introduces the basic structure of a ggplot2 code:

- Start with ggplot() specifying the dataset and aesthetics (aes) mapping.
- Add layers (geom_*) to specify the type of plot (e.g., geom_point() for a scatter plot).
- Use labs() to add labels and title to the plot.

## Create a Histogram for the 'confident' variable

```{r}
ggplot(lexdec, aes(x = confident)) +
  geom_histogram(binwidth = 1, 
                 fill = "skyblue", 
                 color = "black") +
  labs(title = "Histogram of Confidence Levels", 
       x = "Confidence Levels", 
       y = "Frequency")
```

## Create a Boxplot for 'confident' variable

```{r}
ggplot(lexdec, aes(x = yearsStudy, y = sex)) +
  geom_boxplot(fill = "lightgreen", color = "darkgreen") +
  labs(title = "Boxplot of Years of Study by Sex", 
       x = "Sex", 
       y = "Years of Study")
```

## Create a Scatter Plot for 'hoursDay' vs 'yearsStudy'

```{r}
ggplot(lexdec, aes(x = monthsLive, y = RT)) +
  geom_point(color = "blue") +
  labs(title = "Scatter Plot of Hours Study vs RT", 
       x = "Hours Study", 
       y = "Reaction Time (ms)")
```

Most of the data points are squished into the bottom left corner. Let's adjust the scale to zoom in on our plot.

```{r}
ggplot(lexdec, aes(x = monthsLive, y = RT)) +
  geom_point(color = "blue") +
  labs(title = "Scatter Plot of Hours Study vs RT", 
       x = "Hours Study", 
       y = "Reaction Time (ms)") + 
  xlim(0, 500) + # Adjust x-axis limits from 0 to 500
  ylim(0, 10000) # Adjust y-axis limits from 0 to 1e+05
```

Our plot is still not very clear. The integers on the x-axis are causing our data points to overlap. To fix this, include a jitter to move the data points around a little. We can also make the points transparent!

```{r}
ggplot(lexdec, aes(x = monthsLive, y = RT)) +
  # Add a jitter to improve readability
  geom_point(position = position_jitter(width = 0.5, height = 0),
             alpha = 0.3, # Add transparency to data points
             color = "blue") +
  labs(title = "Scatter Plot of Hours Study vs RT", 
       x = "Hours Study", 
       y = "Reaction Time (ms)") + 
  xlim(0, 500) +
  ylim(0, 10000)
```

## Create a Bar Chart for 'domLang'

```{r}
ggplot(lexdec, aes(x = domLang)) +
  geom_bar(fill = "orange") +
  labs(title = "Bar Chart of Dominant Languages", 
       x = "Dominant Languages", 
       y = "Frequency")
```

Our x-axis tick labels are overlapping. We can rotate them to make them readable.

```{r}
ggplot(lexdec, aes(x = domLang)) +
  geom_bar(fill = "orange") +
  labs(title = "Bar Chart of Dominant Languages", 
       x = "Dominant Languages", 
       y = "Frequency") + 
  # Rotate x-axis labels by 45 degrees
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  
```

# Exercises

## Exercise 1: Histogram
Create a histogram to visualize the distribution of 'confident' values in the lexdec dataset.

## Exercise 2: Box Plot
Generate a box plot to compare the 'yearsStudy' variable across different 'domLang' categories in the lexdec dataset.

## Exercise 3: Scatter Plot
Create a scatter plot to explore the relationship between 'hoursDay' and 'RT' (reaction time) variables in the lexdec dataset.

## Exercise 4: Bar Chart
Construct a bar chart to display the frequency of each 'domLang' category in the lexdec dataset.
